GMSE: an R package for generalised management strategy evaluation

### 2017-11-09 notebook/ms/SI1.Rmd
More edits

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above). Manager utility can therefore change in each GMSE time step as resource abundance (and its estimate from the observation model) changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive. To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, and the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,

@@

The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy is not predicted to get closer to the `manage_target` abundance. Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting $D_{i}$ from this value for each individual,

@@

We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used). We now explain in more detail how individuals in the genetic algorithm calculate how their actions will affect $\Delta A_{i}$. These calculations apply to all individuals identically, so we note that the subscript $i$ is dropped; instead, the subscript $j$ refers to actions (e.g., scaring, culling, etc.).

To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ performed by users in the previous time step ($X_{j}^{old}$), and the cost of performing each action ($C^{old}_{j}$). This information is collected from `ACTION` and `COST` arrays. The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{j}^{new}$). To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy. The predicted total number of a particular action $j$ performed is thereby calculated as,

$$ X_{j}^{new} =  (X_{j}^{old} + 1) \frac{C^{old}_{j}}{C^{new}_{j}}. $$

The variable $C^{new}_{j}$ defines the new cost set by the individual for action $j$. A value of 1 is added to ($X_{j}^{old}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step. Otherwise, if $X_{j}^{old} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.

The predicted consequences of $X^{new}_{j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`). For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to greatly overestimate how much the actions attempted by users will actually translate to a change in resource abundance. In practice, the default $\sigma = 0.1$ performs well. Allowed actions are predicted by managers to have the following effects:

@@

These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions. The summation of $X^{new}_{j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.

@@

$$F_{i}^{u} = \Delta A_{i} U^{u}_{res} + \Delta L_{i} U^{u}_{land} $$.

Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively. Further, in GMSE v0.3.1.9, only one term of the equation is nonzero. When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = -1$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield). Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `resource_consume` is greater than zero.

@@

Actions on resources can also have indirect affects on $\Delta L_{i}$ when resources consume output on the landscape (i.e., when `res_consume` is greater than zero), which we define as $r$. The predicted $\Delta L_{i}$ is therefore,

@@

The fitness functions defined above are useful heuristics for simulating manager and user decision-making in a way that produces realistic, *I know it when I see it*, strategies. Future versions of GMSE might improve upon these heuristics to generate more accurate or more realistic models of human decision making. Such improvements could incorporate additional information such as memory of actions from multiple past time steps, or a continually updated estimate for how actions are predicted to affect resource abundance in a simulation (e.g., through a dynamic `manager_sense`). Alternatively, future improvements could usefully incorporate knowledge of human decision making collected from empirical observation of human behaviour during conservation conflicts. While such possibilities could be useful for future GMSE modelling, repeated simulations demonstrate the ability of the current GMSE genetic algorithm to find adaptive strategies for managers attempting to keep resources at target abundance, and users attempting to maximise their harvests or crop yields. It should therefore be useful as a tool for modelling manager and user decisions in a generalised management strategy evaluation framework.

>>>>

Individual fitness as calculated for managers ($F_{i}^{m}$) is affected by a manager's utility for resources and the projected change in resource abundance caused by the individual's policy (i.e., the contents of their `POPULATION` layer, specifically row 3). Manager utility for a resource ($U^{m}_{res}$) is defined as the difference between `manage_target` and the estimation of population abundance as produced by the GMSE observation model (see "[Key data structures used](#Key_data_structures_used)" above). Manager utility can therefore change in each GMSE time step as estimated resource abundance changes; when the estimated resource abundance is greater than `manage_target`, $U^{m}_{res}$ is negative, and when the estimated resource abundance is less than `manage_target`, $U^{m}_{res}$ is positive. To get individual fitness, first the change in resource abundance predicted by the individual's policy ($\Delta A_{i}$) is calculated, then the squared difference between $\Delta A_{i}$ and $U^{m}_{res}$ is calculated to obtain a utility deviation ($D_{i}$) for the individual $i$,

@@

The value of $D_{i}$ increases as $\Delta A_{i}$ gets further from $U^{m}_{res}$; i.e, $D_{i}$ is high when $i$ sets a policy that is not predicted to get closer to the `manage_target` abundance. Fitness is defined by first finding the maximum $D_{i}$ value among all `ga_popsize` individuals ($D_{max}$), then subtracting $D_{i}$ from this value for each individual,

@@

We have explained how $U^{m}_{res}$ is calculated in the [above section on key data structures](#Key_data_structures_used). We now explain in more detail how individuals in the genetic algorithm calculate how their actions will affect $\Delta A_{i}$.

To predict change resource abundance as a consequence of policy, an individual first needs to know the total number of actions of all types $j$ (e.g., scaring, culling, etc.) performed by users in the previous time step ($X_{\bullet, j}$), and the cost of performing each action ($C_{\bullet, j}$). This information is collected from `ACTION` and `COST` arrays. The individual then needs to predict how their policy (i.e., the costs that they set for users to perform an action) will affect the new total number of each action performed ($X_{i,j}$). To do this, the individual assumes that total user actions performed under their policy will change in proportion to that of the old policy. The predicted total number of a particular action $j$ performed under the policy of $i$ is thereby calculated as,

$$ X_{i,j} =  (X_{\bullet, j} + 1) \frac{C_{\bullet, j}}{C_{i,j}}. $$

The variable $C_{i,j}$ defines the new cost set by the individual $i$ for action $j$. A value of 1 is added to ($X_{\bullet, j}$) to model some degree of caution by the manager (this can be changed from the default 1 using `manage_caution`), especially so that managers do not na&#0239;vely assume that users will not perform an action just because they did not perform it in the previous time step. Otherwise, if $X_{\bullet, j} = 0$, then the manager would always assume that a change in the cost of an action would have no effect on the number of times the action was performed by users; a value of 1 assumes that at least one user will perform the action in the new time step.

The predicted consequences of $X_{i,j}$ for resource abundance differ for each possible action. For each action, no consequence is predicted if the policy is not allowed by a simulation of GMSE (e.g., `culling = FALSE`). For allowed actions, the parameter `manager_sense` ($\sigma$) modulates predicted consequences for abundance by some factor; this is useful because not all actions attempted by users will be realised, and a value of $\sigma = 1$ tends to greatly overestimate how much the actions attempted by users will actually translate to a change in resource abundance. In practice, the default $\sigma = 0.1$ performs well. Allowed actions are predicted by managers to have the following effects:

@@

These effects cannot be altered directly in `gmse` or `gmse_apply` (though parameter values can of course be changed using `manager_sense` and `lambda` arguments), but future versions of GMSE might include different predicted effects to increase precision or allow for multiple resource types or different actions. The summation of $X_{i,j}$ for all actions defines the predicted change in resource abundance caused by the policy of an individual $i$, $\Delta A_{i}$.

@@

$$F_{i}^{u} = \Delta A_{i} U^{u}_{res} + \Delta L_{i} U^{u}_{land}. $$

Note that $F_{i}^{u}$ increases when $\Delta A_{i}$ and $\Delta L_{i}$ are of the same sign as $U^{u}_{res}$ and $U^{u}_{land}$, respectively. Further, in GMSE v0.3.1.9, only one term of the equation is nonzero. When `land_ownership = FALSE` (default, modelling users that harvest resources), $U^{u}_{res} = -1$ and $U^{u}_{land} = 0$, and when `land_ownership = TRUE`, $U^{u}_{res} = 0$ and $U^{u}_{land} = 100$ (modelling farmers trying to increase crop yield). Hence users only have a single objective of either decreasing resource abundance or increasing landscape output, though landscape output might be increased indirectly by decreasing resource abundance if `resource_consume` is greater than zero.

@@

Actions on resources can also have indirect effects on $\Delta L_{i}$ when resources consume output on the landscape; we define the value `res_consume` as $r$. The predicted $\Delta L_{i}$ is then,

@@

The fitness functions defined above are useful heuristics for simulating manager and user decision-making in a way that produces realistic, *I know it when I see it*, strategies. Future versions of GMSE might improve upon these heuristics to generate more accurate or more realistic models of human decision making. Such improvements could incorporate additional information such as memory of actions from multiple past time steps, or a continually updated estimate for how actions are predicted to affect resource abundance or landscape output in a simulation (e.g., through a dynamic `manager_sense`). Alternatively, future improvements could usefully incorporate knowledge of human decision making collected from empirical observation of human behaviour during conservation conflicts. While such possibilities could be useful for future GMSE modelling, repeated simulations demonstrate the ability of the current GMSE genetic algorithm to find adaptive strategies for managers attempting to keep resources at target abundance, and users attempting to maximise their harvests or crop yields. It is therefore useful as a tool for modelling manager and user decisions in a generalised management strategy evaluation framework.
